---
output:
  rmarkdown::github_document:
    html_preview: false
---

# Software Development Standards ![GitHub last commit](https://img.shields.io/github/last-commit/USCbiostats/software-dev)

This project's main contents are located in the project's [Wiki](https://github.com/USCbiostats/CodingStandards/wiki).

If you want an account on **Bioghost**, fillout the form [here](https://forms.gle/GLKGfydPj5LmfHWq6) (USC accout required).

# USCbiostats R packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

As of `r Sys.Date()`, the packages listed here have been cited `r tot_citations` times
(source: Google Scholar).


```{r listing-pkgs, echo = FALSE}
pkgs <- read.csv("packages.csv", comment.char = "#", stringsAsFactors = FALSE)

# Alphabetically ordered
pkgs <- pkgs[order(pkgs$name),,drop=FALSE]

dat <- with(pkgs, data.frame(
  Name        = sprintf(
    "[**%s**](%s)", name, ifelse(!is.na(repo) & repo != "", repo, paste0("https://github.com/USCbiostats/", name))
    ),
  Description = paste(
    description,
    sprintf(
      "[![CRAN status](https://www.r-pkg.org/badges/version/%s)](https://CRAN.R-project.org/package=%1$s)",
      name
    ),
    sprintf(
      "[![CRAN downloads](http://cranlogs.r-pkg.org/badges/grand-total/%s)](https://cran.r-project.org/package=%1$s)",
      name
    ),
    sprintf(
      "[![status](https://tinyverse.netlify.com/badge/%s)](https://CRAN.R-project.org/package=%1$s)",
      name
    )
), stringsAsFactors = FALSE))

for (pkg in pkgs$name[which(pkgs$on_bioc)]) {
  dat[which(pkgs$name == pkg), "Description"] <- 
  paste0(pkgs$description[pkgs$name == pkg], 
  sprintf("[![](https://img.shields.io/badge/Bioconductor%%20version-1.0.0-green.svg)](https://www.bioconductor.org/packages/%s)", pkg), 
  badger::badge_bioc_download(pkg, "total", "blue", "total"))
}

```

```{r citations, include=FALSE}
regex <- "([0-9,]+)[\\s\\n]+results?[\\s\\n]+\\([\\s\\n]*[0-9]+" #[\\s\\n]+[(][0-9]*[.]?[0-9]+\\s+secs?
dat$Citations <- ""
tot_citations <- 0L
for (i in seq_len(nrow(pkgs))) {
  
  # If no URL, then continue
  if (nchar(pkgs$google_scholar[i]) == 0) {
    next
  }
  
  # Otherwise, take a look at the cictations
  address <- pkgs$google_scholar[i]
  
  page      <- tryCatch(readLines(address, warn = FALSE), error = function(e) e)

  if (inherits(page, "error"))
    next
  
  # Removing blocks of bold, italic, etc
  page <- gsub("\\<[[:alnum:]_/-]+\\>", "",page, perl = TRUE)
  
  citations <- which(grepl(regex, page, perl = TRUE))
  if (!length(citations))
    next
  
  citations <- stringr::str_extract(page[citations], "[0-9,]+(?=[\\s\\n]+results?)")
  citations <- as.integer(gsub("[,.]", "", citations, perl = TRUE))
  tot_citations <- tot_citations + citations
  dat$Citations[i] <- sprintf("[%i](%s)", as.integer(citations), address)
  
}
```


```{r checking-urls, echo = FALSE, eval = FALSE}
http_status <- sapply(dat$Name, function(n) {
  message("Checking ", n)
  web <- httr::GET(
    sprintf("https://github.com/USCbiostats/%s", n),
    httr::config(connecttimeout = 60)
    )
  
  httr::status_code(web)
})

table(http_status) # 22
dat$Name <- sprintf("[**%s**](https://github.com/USCbiostats/%s)", dat$Name, dat$Name)
```


```{r printing, echo = FALSE}
knitr::kable(dat, row.names = FALSE)
```

To update this list, modify the file [packages.csv](packages.csv). The
`README.md` file is updated automatically using GitHub Actions, so there's no
need to "manually" recompile the README file after updating the list. 

# Coding Standards

1.  [Coding Standards](../../wiki/Home#coding-standards)
2.  [Software Thinking](../../wiki/Home#software-thinking)
3.  [Development Workflow](../../wiki/Home#development-workflow)
4.  [Misc](../../wiki/Home#misc)

We do have some direct guidelines developed as issue templates [here](templates). 

# Bioghost Server

1.  [Introduction](../../wiki/Biogohst-server#introduction)
2.  [Setup](../../wiki/Biogohst-server#setup)
3.  [Cheat Sheet](../../wiki/Biogohst-server#cheat-sheet)

# HPC in R
    
*   [Parallel computing in R](../../wiki/HPC-in-R#parallel-computing-in-r)  
*   [parallel](../../wiki/HPC-in-R#parallel)
*   [iterators+foreach](../../wiki/HPC-in-R#foreach)
*   [RcppArmadillo + OpenMP](../../wiki/HPC-in-R#rcpparmadillo-and-openmp)

# Happy Scientist Seminars

The Happy Scientist Seminars are educational seminars sponsored by Cores C and D of IMAGE, the Biostats Program Project award. This series, the "Happy Scientist" seminar series, is aimed at providing educational material for members of Biostats, both students and faculty, about a variety of tools and methods that might prove useful to them. If you have any suggestions for subjects that you would like to learn about in future, please send email to Paul Marjoram at (pmarjora@usc.edu). Our agenda will be driven by your specific interests as far as is possible. 

List of past seminars with material can be found [here](/happy_scientist/).
