---
title: "Yet Another 'Writing R packages' tutorial"
author: 'George G. Vega Yon<br><a href="http://cana.usc.edu/vegayon">cana.usc.edu/vegayon</a><br><a href="mailto:vegayon@usc.edu">vegayon@usc.edu</a>'
date: "January 24, 2017"
output: 
  slidy_presentation:
    footer: vegayon@usc.edu
    incremental: false
    font_adjustment: -1
    duration: 45
    css: ggvy_style.css
    slide_level: 1
---

```{r, echo = FALSE}
knitr::opts_chunk$set(out.width = "600px", fig.align = "center")
```


# Today

*   Have a look at some R-package-dev practices based mostly on my own experience (so the product is sold AS-IS)
*   We won't discuss either Rcpp or coding R itself.
*   We won't discuss github in depth. I'll assume that the user already knows the basics of git(hub): Creating a repo online, and start using it locally.
*   We won't talk about software engineering/design, rather we'll take a look at some tools and dev cycle.

# What we'll be using today

a.  **devtools**: "Package development tools for R" ([read more](https://cran.r-project.org/package=devtools)).
    
b.  **roxygen2**: "A 'Doxygen'-like in-source documentation system for Rd, collation, and 'NAMESPACE' files"
    ([read more](https://cran.r-project.org/package=roxygen2)).

c.  **covr**: (An R package for) "Test Coverage for Packages" ([read more](https://cran.r-project.org/package=covr)).

d.  **RStudio**: "RStudio is an integrated development environment (IDE) for R" ([read more](https://cran.r-project.org/package=covr)).

c.  **valgrind**: (if you use C/C++ code) "a memory error detector" ([read more](http://valgrind.org/)).


# Why 'spending' time writing an R package?

To name a few:

1.  Easiest way to share code: Just type `install.packages` and voilà!

2.  Already standarized: So you don't need to think about how to structure it.

3.  CRAN checks everything for you: Force yourself to code things right.

4.  ...


# What's an R package, anyway?

According to Hadley Wickham's ["R packages"](http://r-pkgs.had.co.nz/)

> Packages are **the fundamental units of reproducible R code**. They include
**reusable R functions**, the **documentation** that describes how to use them, and
**sample data**.



# Dev cycle

Writing R packages is an iterative process

1.  **Set up the structure**: Create folders and files `R/`, `data/`, `src/`, `tests/`, `man/`,
    `vignettes/`, `DESCRIPTION`, `NAMESPACE`(?)

2.  **Code!** For `f` in `F` do
    
    1.  Write `f`
    
    2.  Document `f`: what it does, what the inputs are, what it returns, examples.
    
    3.  Add some tests: is it doing what is supposed to do?
    
    4.  `R CMD Check`: Will CRAN, and the rest of the functions, 'like' it?
    
    5.  Commit your changes!: Update `ChangeLog`+`news.md`, and commit the changes!
    
    6.  next `f`


# Step 1: Set up the structure

You have several methods: package.skeleton(), Rstudio's "New Project", etc. I
personally do it from scratch (including the code):

1.  Create the package dir, say `funnypkg`

2.  Create the `R` dir

3.  Create the `DESCRIPTION` file ([read more](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#The-DESCRIPTION-file)):
    
    ```
    Package: funnypkg
    Type: Package
    Title: What the Package Does (Title Case)
    Version: 0.1.0
    Author: Who wrote it
    Maintainer: The package maintainer <yourself@somewhere.net>
    Description: More about what it does (maybe more than one line)
        Use four spaces when indenting paragraphs within the Description.
    License: MIT + file
    Encoding: UTF-8
    LazyData: true
    ```

4.  From RStudio, create a new project pointing to the package
    folder.

----

## Step 1: Set up the structure (cont. 1)

And, using the `devtools` package, we can add some extras

```r
# Creating a README for the project
devtools::use_readme_rmd()

# Infrastructure for testing
devtools::use_testthat()

# Infrastructure for Code Coverage
devtools::use_coverage(type = "codecov") # This creates the .travis.yml
devtools::use_appveyor()

# A LICENSE file (required by CRAN)
# The line "license: MIT + file" in the DESCRIPTION file
devtools::use_mit_license(copyright_holder = "George G. Vega Yon")
devtools::use_news_md()
```

----

## Step 1: Set up the structure (cont. 2)

What did just happen?

1.  `use_readme_rmd()`: Creates a readme file that will be in the main folder
    of the project. Think about it as the home page of your project.
    
2.  `use_testthat()`: Will create the basic infrastructure for package testing.

3.  `use_coverage(type = "codecov")`: Creates the [.travis.yml file](https://en.wikipedia.org/wiki/Comparison_of_continuous_integration_software)
    for Unix CI, and
    sets it up for code coverage using [codecov.io](https://codecov.io/).
    
4.  `devtools::use_appveyor()`: Creates the [appveyor.yml](https://www.appveyor.com/)
    for windows CI.
    
5.  `devtools::use_mit_license(copyright_holder = "George G. Vega Yon")`: Creates the----
    `LICENSE` file and puts it under 'George G. Vega Yon'.

6.  `devtools::use_news_md()`: Creates the `news.md` file which us used for
    tracking changes, and communicating them to the users (e.g.
    [netdiffuseR](https://github.com/USCCANA/netdiffuseR/blob/master/NEWS.md))


# Step 2: Code!
    
## Step 2.1 and 2.2: Write `f` and Document it

*   Using `roxygen2` is very straight forward
        
    ```{r}
    #' The title of -addnums-
    #'
    #' Here is a brief description
    #' 
    #' @param a Numeric scalar. A brief description.
    #' @param b Numeric scalar. A brief description.
    #' 
    #' @details Computes the sum of \code{x} and \code{y}.
    #' @return A list of class \code{funnypkg_addnums}:
    #' \item{a}{Numeric scalar.}
    #' \item{b}{Numeric scalar.}
    #' \item{ab}{Numeric scalar. the sum of \code{a} and \code{b}}
    #' @examples
    #' foo(1, 2)
    #' 
    #' @export
    addnums <- function(a, b) {
      ans <- a + b
      structure(list(a = a, b = b, ab = ans)
      , class = "funnypkg_addnums")
    }
    
    ```

*   Press Ctrl + Shift + D (RStudio will: create the manual, and the NAMESPACE).
    Make sure you activate this option in RStudio (not the default).
    
*   Notice that the output is defined using S3-type objects ([read more](http://adv-r.had.co.nz/S3.html))

---

## Step 2.1 and 2.2: Write `f` and Document it (cont. 1)

*   Here we are adding a function that is documented in `addnums`. The plot method
    
    ```{r}
    #' @rdname addnums
    #' @export
    #' @param x An object of class \code{funnypkg_addnums}.
    #' @param y Ignored.
    #' @param ... Further arguments passed to
    #' \code{\link[graphics:plot.window]{plot.window}}.
    plot.funnypkg_addnums <- function(x, y = NULL, ...) {
      graphics::plot.new()
      graphics::plot.window(xlim = range(unlist(c(0,x))), ylim = c(-.5,1))
      graphics::axis(1)
      with(x, graphics::segments(0, 1, ab, col = "blue", lwd=3))
      with(x, graphics::segments(0, 0, a, col = "green", lwd=3))
      with(x, graphics::segments(a, .5, a + b, col = "red", lwd=3))
      graphics::legend("bottom", col = c("blue", "green", "red"), 
             legend = c("a+b", "a", "b"), bty = "n", 
             ncol = 3, lty = 1, lwd=3)
    }
    
    ```
      
*   You'll need to update the `man/*Rd` everytime that you add new roxygen content.
    Just press Ctrl + Shift + D and RStudio will do it for you.
    
*   Notice that functions from foreign packages are called using the :: operator
    ([read more](https://stat.ethz.ch/R-manual/R-devel/library/base/html/ns-dblcolon.html)). Not a requirement, but makes the code easier to maintain.

---

## Step 2.1 and 2.2: Write `f` and Document it (cont. 2)

Since we added foreign functions, we need to add them to the NAMESPACE and
to the DESCRIPTION files:

*   For the NAMESPACE, we can use `roxygen` as follows:
    
    ```r
    #' @importFrom graphics plot.new plot.window axis legend
    NULL
    
    #' funnypkg
    #'
    #' A (not so) funny collection of functions
    #'
    #' @description We add stuff up... You can access to the project
    #' website at \url{http://github.com/USCBiostats/funnypkg}
    #'
    #' @docType package
    #' @name funnypkg
    #'
    #' @author George G. Vega Yon
    NULL
    ```
    
*   Here is the DESCRIPTION updated
    
    ```
    Package: funnypkg
    Type: Package
    Title: What the Package Does (Title Case)
    Version: 0.1.0
    Author: Who wrote it
    Maintainer: The package maintainer <yourself@somewhere.net>
    Description: More about what it does (maybe more than one line)
        Use four spaces when indenting paragraphs within the Description.
    License: What license is it under?
    Encoding: UTF-8
    LazyData: true
    Imports: graphics
    ```

# Step 3: Add some tests

In the `tests/testthat/` dir, add/edit a source file with tests, e.g. `test-basic.r`

```r
context("Basic set of tests")
test_that("foo(a, b) = a+b", {
  # Preparing the test
  a <- 1
  b <- -2
  
  # Calling the function
  ans0 <- a+b
  ans1 <- foo(a, b)
  
  # Are these equal?
  expect_equal(ans0, ans1$ab)
})

test_that("Plot returns -funnypkg_foo-", {
  expect_s3_class(plot(foo(1,2)), "funnypkg_foo")
})
```

You can run the tests using Ctrl + Shift + t

# Step 4: Checking the package

1.  If you don't have C/C++  code Just press Ctrl + Shift + E

2.  If you have C/C++ code, use `R CMD Check` with valgrind (check for segfaults)

    ```shell
    $ R CMD build funnypkg
    $ R CMD check --as-cran --use-valgrind funnypkg*.tar.gz
    ```

    You can ask RStudio to use valgrind too.


# Step 5: Commit your Changes

*   If after adding/changing the code nothing breaks, then you are good to go with your changes!
    
    ```shell
    $ git commit -a -m "Adding function abc"
    $ git pull
    $ git push
    ```

*   Ideally, you'll want to track changes using `ChangeLog` and `NEWS.md` files.

# References

Mostly from experience, and

*   Hadley's R-pkgs
*   Writing R Extensions
