---
title: "Happy Scientist:<br>Yet Another 'Writing R packages' tutorial"
author: 'George G. Vega Yon<br><a href="http://cana.usc.edu/vegayon">cana.usc.edu/vegayon</a><br><a href="mailto:vegayon@usc.edu">vegayon@usc.edu</a>'
date: "February 9, 2017"
output: 
  slidy_presentation:
    footer: vegayon@usc.edu
    incremental: false
    font_adjustment: -1
    duration: 45
    css: ggvy_style.css
    slide_level: 2
    highlight: haddock
---

```{r, echo = FALSE}
knitr::opts_chunk$set(out.width = "600px", fig.align = "center")
```


# Today

*   Have a look at some R-package-dev practices based mostly on my own experience
    (so the product is sold AS-IS)
    
*   We won't discuss either Rcpp or coding R itself. (for that I
    recommend ["Advance R"](http://adv-r.had.co.nz/Rcpp.html), and there's also
    Patrick Muchmore's ["A template for R packages using Rcpp and/or RcppArmadillo "](https://github.com/patrickmuchmore/RPackageTemplate))

*   We won't discuss github in depth. I'll assume that the user already knows the
    basics of git(hub): Creating a repo online, and start using it locally.
    (A good reference is ["Pro Git"](https://git-scm.com/book/en/v2))
    
*   We won't talk about software engineering/design, rather we'll take a look at
    some tools and dev cycle. (For that you have ["Code Complete"](http://cc2e.com/))
    

# What we'll be using today

a.  **devtools**: "Package development tools for R" ([read more](https://cran.r-project.org/package=devtools)).
    
b.  **roxygen2**: "A 'Doxygen'-like in-source documentation system for Rd, collation, and 'NAMESPACE' files"
    ([read more](https://cran.r-project.org/package=roxygen2)).

c.  **covr**: (An R package for) "Test Coverage for Packages" ([read more](https://cran.r-project.org/package=covr)).

d.  **RStudio**: "RStudio is an integrated development environment (IDE) for R" ([read more](https://cran.r-project.org/package=covr)).

c.  **valgrind**: (if you use C/C++ code) "a memory error detector" ([read more](http://valgrind.org/)).


# Why 'spending' time writing an R package?

To name a few:

1.  Easy way to share code: Just type `install.packages` and voilà!

2.  Already standarized: So you don't need to think about how to structure it.

3.  CRAN checks everything for you: Force yourself to code things right.

4.  **Reproducible Research**
    
    *   Accessible Reproducible Research ([Mesirov, __Science__ 2010](http://science.sciencemag.org/content/327/5964/415))
    
    *   Again, and Again, and Again, ... ([Jasny et al., __Science__ 2011](http://science.sciencemag.org/content/334/6060/1225))
        
    *   Challenges in Irreproducible Research ([__nature__ topic](http://www.nature.com/news/reproducibility-1.17552))


# What's an R package, anyway?

According to Hadley Wickham's ["R packages"](http://r-pkgs.had.co.nz/)

> Packages are **the fundamental units of reproducible R code**. They include
**reusable R functions**, the **documentation** that describes how to use them, and
**sample data**.

```{r What it has, echo=FALSE}
words <- c("R/", "tests/","vignettes/", "data/", "man/",
           "DESCRIPTION", "NAMESPACE")
freqs <- 7:1
# oldpar <- par(no.readonly = TRUE)
# par(mar = rep(0,4))
# wordcloud::wordcloud(words, colors = c(blues9[5:9], gray(c(.8, .6)))[7:1],
#                      freqs, min.freq = 1)
# par(oldpar)

library(d3wordcloud)
d3wordcloud(words, freqs, colors = c(blues9[5:9], gray(c(.8, .6))))
```

# What's an R package, anyway? (cont.)

Folders

*   **R** Where the R scripts live, e.g. `addnums.r`, `funny-pkg.r`.

*   **tests** Where the tests live, e.g. `testthat/test-addnums.r` (using testthat).

*   **vignettes** Where vignettes live, e.g. `vignettes/mode_details_on_addnums.rmd`.

*   **data** Where the data lives, e.g. `fakedata.rda`, `aneattable.csv`.

*   **man** Where the manuals live, e.g. `addnums.Rd`, `funnypkg.Rd`.

And files

*   **DESCRIPTION** The metadata of the package: name, description, version, author(s),
    dependencies, etc.

*   **NAMESPACE** Declares the name of the objects (functions mostly) that will
    be part of the package (and will be visible to the user)

More info [here](http://r-pkgs.had.co.nz/package.html) and [here](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Package-structure)


# Dev cycle

Writing R packages is an iterative process

1.  **Set up the structure**: Create folders and files `R/`, `data/`, `src/`, `tests/`, `man/`,
    `vignettes/`, `DESCRIPTION`, `NAMESPACE`(?)

2.  **Code!** For `f` in `F` do
    
    1.  Write `f`
    
    2.  Document `f`: what it does, what the inputs are, what it returns, examples.
    
    3.  Add some tests: is it doing what is supposed to do?
    
    4.  `R CMD Check`: Will CRAN, and the rest of the functions, 'like' it?
    
    5.  Commit your changes!: Update `ChangeLog`+`news.md`, and commit
        (so travis and friends run)!
    
    6.  next `f`
    
3.  Submit your package to CRAN/Bioconductor (read [more](http://r-pkgs.had.co.nz/release.html#release-submission))


# Step 1: Set up the structure

You have several methods: `package.skeleton()`, Rstudio's "New Project", `devtools::new()`, etc. I
personally do it from scratch (including the code):

1.  Create the package dir, say `funnypkg`

2.  Create the `R` dir

3.  Create the `DESCRIPTION` file ([read more](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#The-DESCRIPTION-file)):
    
    ```
    Package: funnypkg
    Type: Package
    Title: What the Package Does (Title Case)
    Version: 0.1.0
    Author: Who wrote it
    Maintainer: The package maintainer <yourself@somewhere.net>
    Description: More about what it does (maybe more than one line)
        Use four spaces when indenting paragraphs within the Description.
    License: MIT + file
    Encoding: UTF-8
    LazyData: true
    ```

4.  From RStudio, create a new project pointing to the package
    folder.

----

## Step 1: Set up the structure (cont. 1)

And, using the `devtools` package, we can add some extras

```r
# Creating a README for the project
devtools::use_readme_rmd()

# Infrastructure for testing
devtools::use_testthat()

# Infrastructure for Code Coverage
devtools::use_coverage(type = "codecov") # This creates the .travis.yml
devtools::use_appveyor()

# A LICENSE file (required by CRAN)
# The line "license: MIT + file" in the DESCRIPTION file
devtools::use_mit_license(copyright_holder = "George G. Vega Yon")
devtools::use_news_md()
```

----

## Step 1: Set up the structure (cont. 2)

What did just happen?

1.  `use_readme_rmd()`: Creates a readme file that will be in the main folder
    of the project. Think about it as the home page of your project.
    
2.  `use_testthat()`: Will create the basic infrastructure for package testing.

3.  `use_coverage(type = "codecov")`: Creates the [.travis.yml file](https://en.wikipedia.org/wiki/Comparison_of_continuous_integration_software)
    for Unix CI, and
    sets it up for code coverage using [codecov.io](https://codecov.io/).
    
4.  `devtools::use_appveyor()`: Creates the [appveyor.yml](https://www.appveyor.com/)
    for windows CI.
    
5.  `devtools::use_mit_license(copyright_holder = "George G. Vega Yon")`: Creates the----
    `LICENSE` file and puts it under 'George G. Vega Yon'.

6.  `devtools::use_news_md()`: Creates the `news.md` file which us used for
    tracking changes, and communicating them to the users (e.g.
    [netdiffuseR](https://github.com/USCCANA/netdiffuseR/blob/master/NEWS.md))


# Step 2: Code!
    
## Step 2.1 and 2.2: Write `f` and Document it

*   Using `roxygen2` is very straight forward. For our fist pice of code, we create the
    `R/addnums.r` file:
        
    ```{r}
    #' The title of -addnums-
    #'
    #' Here is a brief description
    #' 
    #' @param a Numeric scalar. A brief description.
    #' @param b Numeric scalar. A brief description.
    #' 
    #' @details Computes the sum of \code{x} and \code{y}.
    #' @return A list of class \code{funnypkg_addnums}:
    #' \item{a}{Numeric scalar.}
    #' \item{b}{Numeric scalar.}
    #' \item{ab}{Numeric scalar. the sum of \code{a} and \code{b}}
    #' @examples
    #' foo(1, 2)
    #' 
    #' @export
    addnums <- function(a, b) {
      ans <- a + b
      structure(list(a = a, b = b, ab = ans)
      , class = "funnypkg_addnums")
    }
    
    ```

*   Type `devtools::document()`, or press Ctrl + Shift + D (RStudio will: create the manual, and the NAMESPACE).
    Make sure you activate this option in RStudio (not the default).
    
*   Notice that the output is defined using S3-type objects ([read more](http://adv-r.had.co.nz/S3.html))

---

## Step 2.1 and 2.2: Write `f` and Document it (cont. 1)

*   Here we are adding a function that is documented in `addnums`. The plot method
    
    ```{r}
    #' @rdname addnums
    #' @export
    #' @param x An object of class \code{funnypkg_addnums}.
    #' @param y Ignored.
    #' @param ... Further arguments passed to
    #' \code{\link[graphics:plot.window]{plot.window}}.
    plot.funnypkg_addnums <- function(x, y = NULL, ...) {
      graphics::plot.new()
      graphics::plot.window(xlim = range(unlist(c(0,x))), ylim = c(-.5,1))
      graphics::axis(1)
      with(x, graphics::segments(0, 1, ab, col = "blue", lwd=3))
      with(x, graphics::segments(0, 0, a, col = "green", lwd=3))
      with(x, graphics::segments(a, .5, a + b, col = "red", lwd=3))
      graphics::legend("bottom", col = c("blue", "green", "red"), 
             legend = c("a+b", "a", "b"), bty = "n", 
             ncol = 3, lty = 1, lwd=3)
    }
    
    ```
      
*   You'll need to update the `man/*Rd` every time that you add new roxygen content.
    Just press Ctrl + Shift + D and RStudio will do it for you.
    
*   Notice that functions from foreign packages are called using the `::` operator
    (read more [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/ns-dblcolon.html)
    and [here](http://r-pkgs.had.co.nz/namespace.html)). Not a requirement, but
    makes the code easier to maintain (and less error-prone... because of R's scoping rules).

---

## Step 2.1 and 2.2: Write `f` and Document it (cont. 2)

Since we added foreign functions, we need to add them to the NAMESPACE and
to the DESCRIPTION files:

*   Here is the `R/funny-pkg.r` file:
    
    ```r
    #' @importFrom graphics plot.new plot.window axis legend
    NULL
    
    #' funnypkg
    #'
    #' A (not so) funny collection of functions
    #'
    #' @description We add stuff up... You can access to the project
    #' website at \url{https://github.com/USCbiostats/software-dev/tree/master/rpkgs/funnypkg}
    #'
    #' @docType package
    #' @name funnypkg
    #'
    #' @author George G. Vega Yon
    NULL
    ```
    
*   Here is the DESCRIPTION updated
    
    ```
    Package: funnypkg
    Type: Package
    Title: What the Package Does (Title Case)
    Version: 0.1.0
    Author: Who wrote it
    Maintainer: The package maintainer <yourself@somewhere.net>
    Description: More about what it does (maybe more than one line)
        Use four spaces when indenting paragraphs within the Description.
    License: What license is it under?
    Encoding: UTF-8
    LazyData: true
    Suggests:
        testthat,
        covr
    RoxygenNote: 5.0.1
    Imports: graphics
    ```
    
*   The `Suggests` and `RoxygenNote` fields where added automagically by `devtools`.

# Step 2.3: Add some tests

Test are run every time that you run R CMD check

*   In the `tests/testthat/` dir, add/edit a source file with tests, e.g. `test-addnums.r`
    
    ```r
    context("Basic set of tests")
    test_that("addnums(a, b) = a+b", {
      # Preparing the test
      a <- 1
      b <- -2
      
      # Calling the function
      ans0 <- a+b
      ans1 <- addnums(a, b)
      
      # Are these equal?
      expect_equal(ans0, ans1$ab)
    })
    
    test_that("Plot returns -funnypkg_foo-", {
      expect_s3_class(plot(addnums(1,2)), "funnypkg_foo")
    })
    ```
    
*   You can run the tests using Ctrl + Shift + t

*   Furthermore, tests are use to evaluate code coverage.

# Step 2.4: R CMD check

Running `R CMD check` is easy with RStudio

*   If you don't have C/C++  code Just press Ctrl + Shift + E

*   If you have C/C++ code, use `R CMD Check` with valgrind (check for segfaults)
    
    ```shell
    $ R CMD build funnypkg/
    $ R CMD check --as-cran --use-valgrind funnypkg*.tar.gz
    ```
    
    You can ask RStudio to use valgrind too.

For an extensive list of the checks, see the section
[1.3.1 Checking packages](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Checking-packages)
in the "Writing R extensions" manual.

---

How does the output of `R CMD check` looks like?

```shell
==> devtools::check()

Updating funnypkg documentation
Loading funnypkg
Setting env vars ---------------------------------------------------------------
CFLAGS  : -Wall -pedantic
CXXFLAGS: -Wall -pedantic
Building funnypkg --------------------------------------------------------------
'/usr/lib/R/bin/R' --no-site-file --no-environ --no-save --no-restore --quiet  \
  CMD build  \
  '/home/vegayon/Dropbox/usc/core_project/CodingStandards/rpkgs/funnypkg'  \
  --no-resave-data --no-manual 

* checking for file ‘/home/vegayon/Dropbox/usc/core_project/CodingStandards/rpkgs/funnypkg/DESCRIPTION’ ... OK
* preparing ‘funnypkg’:
* checking DESCRIPTION meta-information ... OK
* checking for LF line-endings in source and make files
* checking for empty or unneeded directories
* building ‘funnypkg_0.1.999.tar.gz’

Setting env vars ---------------------------------------------------------------
_R_CHECK_CRAN_INCOMING_USE_ASPELL_: TRUE
_R_CHECK_CRAN_INCOMING_           : FALSE
_R_CHECK_FORCE_SUGGESTS_          : FALSE
Checking funnypkg --------------------------------------------------------------
'/usr/lib/R/bin/R' --no-site-file --no-environ --no-save --no-restore --quiet  \
  CMD check '/tmp/RtmpBYGSto/funnypkg_0.1.999.tar.gz' --as-cran --timings  \
  --no-manual 

* using log directory ‘/home/vegayon/Dropbox/usc/core_project/CodingStandards/rpkgs/funnypkg.Rcheck’
* using R version 3.3.2 (2016-10-31)
* using platform: x86_64-pc-linux-gnu (64-bit)
* using session charset: UTF-8
* using options ‘--no-manual --as-cran’
* checking for file ‘funnypkg/DESCRIPTION’ ... OK
* checking extension type ... Package
* this is package ‘funnypkg’ version ‘0.1.999’
* package encoding: UTF-8
* checking package namespace information ... OK
* checking package dependencies ... OK
* checking if this is a source package ... OK
* checking if there is a namespace ... OK
* checking for executable files ... OK
* checking for hidden files and directories ... OK
* checking for portable file names ... OK
* checking for sufficient/correct file permissions ... OK
* checking whether package ‘funnypkg’ can be installed ... OK
* checking installed package size ... OK
* checking package directory ... OK
* checking DESCRIPTION meta-information ... OK
* checking top-level files ... OK
* checking for left-over files ... OK
* checking index information ... OK
* checking package subdirectories ... OK
* checking R files for non-ASCII characters ... OK
* checking R files for syntax errors ... OK
* checking whether the package can be loaded ... OK
* checking whether the package can be loaded with stated dependencies ... OK
* checking whether the package can be unloaded cleanly ... OK
* checking whether the namespace can be loaded with stated dependencies ... OK
* checking whether the namespace can be unloaded cleanly ... OK
* checking loading without being on the library search path ... OK
* checking dependencies in R code ... OK
* checking S3 generic/method consistency ... OK
* checking replacement functions ... OK
* checking foreign function calls ... OK
* checking R code for possible problems ... OK
* checking Rd files ... OK
* checking Rd metadata ... OK
* checking Rd line widths ... OK
* checking Rd cross-references ... OK
* checking for missing documentation entries ... OK
* checking for code/documentation mismatches ... OK
* checking Rd \usage sections ... OK
* checking Rd contents ... OK
* checking for unstated dependencies in examples ... OK
* checking examples ... OK
* checking for unstated dependencies in ‘tests’ ... OK
* checking tests ...
  Running ‘testthat.R’ OK
* DONE
Status: OK




R CMD check results
0 errors | 0 warnings | 0 notes

R CMD check succeeded
```

# Step 2.5: Commit your changes

Once you have run R CMD checks and no errors or unexpected issues are detected,
you should commit your changes and push them to Github.

1.  Edit the [`NEWS.md`](https://github.com/hadley/dplyr/blob/master/NEWS.md) file and the
    [`ChangeLog`](https://github.com/RcppCore/RcppArmadillo/blob/master/ChangeLog) (if you have one).

2.  Add the new files to the tree, commit your changes, and push them to Github
    
    ```shell
    $ git add R/newfile.r man/newfile.Rd
    $ git commit -a -m "Adding new function"
    $ git push
    ```
    
Once uploaded, I recommend checking Travis and AppVeyor to see if everything
went well.

# Step 3: Submit your package to CRAN/Bioconductor

*   You can release your package using `devtools::release()` function.

*   The devtools package was built to make life easier. Such is it's creator's
    confidence on the `release` function that it comes with a warranty!
    
    > If a devtools bug causes one of the CRAN maintainers to **treat you
      impolitely**, I will personally send you a handwritten apology note. Please
      forward me the email and your address, and I'll get a card in the mail.

# Dev Cycle (once again!)

The most important step

2.  **Code!** For `f` in `F` do
    
    1.  Write `f`
    
    2.  Document `f`: what it does, what the inputs are, what it returns, examples.
    
    3.  Add some tests: is it doing what is supposed to do?
    
    4.  `R CMD Check`: Will CRAN, and the rest of the functions, 'like' it?
    
    5.  Commit your changes!: Update `ChangeLog`+`news.md`, and commit
        (so travis and friends run)!
    
    6.  next `f`
    
    
# Yet Another 'Writing R packages' tutorial

<body>
<p align="center" style="font-size:50px"><b>Thanks!</b></pr><br>
<p align="center">George G. Vega Yon<br><a href="http://cana.usc.edu/vegayon">cana.usc.edu/vegayon</a><br><a href="mailto:vegayon@usc.edu">vegayon@usc.edu</a></p>
</body>

## More resources

*   ["Writing R extensions"](https://cran.r-project.org/doc/manuals/r-release/R-exts.html)
    (R Core Team)

*   ["R packages"](http://r-pkgs.had.co.nz/) (Hadley Wickham)

*   ["Developing Packages with RStudio"](https://support.rstudio.com/hc/en-us/articles/200486488-Developing-Packages-with-RStudio)
    (Josh Paulson)
    
*   ["A template for R packages using Rcpp and/or RcppArmadillo "](https://github.com/patrickmuchmore/RPackageTemplate)
    (Patrick Muchmore)

*   ["R package primer: a minimal tutorial"](http://kbroman.org/pkg_primer/) (Karl Broman)

*   ["Writing an R package from scratch"](https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/)
    (Hilary Parker)

*   ["Instructions for Creating Your Own R Package"](http://web.mit.edu/insong/www/pdf/rpackage_instructions.pdf)
    (Kim _et al._)
