---
output: github_document
---

# Slow R patterns

```{r}
library(magrittr)
```


This document serves as a starting point for code optimizations. It includes possible optimization strategies.

All code here is being benchmarked using the [bench](http://bench.r-lib.org/) package.

## For loops vs apply

Imagine you have a matrix of numbers and you want to compute the row-wise variance. This can be done using a for loop, or using the apply function. You will most likely not see any big speed improvements, but the lines of codes can sometimes be considerably be improved.

```{r}
data <- matrix(rnorm(100000), nrow = 1000)

rowVars <- function(x) {
  x_length <- nrow(x)
  res <- numeric(x_length)
  for (i in seq_len(x_length)) {
    res[i] <- var(x[i, ])
  }
  res
}

bench::mark(
  rowVars(data),
  apply(data, 1, var)
)
```


## Apply Sums

If you are about to use `apply` with `FUN = mean` or `sum` you might take a look at `rowSums`, `colsums`, `rowMeans` and `colMeans`. They will do the same but faster (NaN and NA are handled slightly different to tread lightly). the `.rowSums` is a faster variant that only works on numeric matrices and do not name the result.

```{r}
data <- matrix(rnorm(10000), nrow = 100, ncol = 100)

bench::mark(
  apply(data, 1, sum),
  rowSums(data),
  .rowSums(data, 100, 100)
) %>% plot()

data <- matrix(rnorm(1000000), nrow = 1000, ncol = 1000)

bench::mark(
  apply(data, 1, sum),
  rowSums(data),
  .rowSums(data, 1000, 1000)
) %>% plot()

bench::mark(
  apply(data, 1, mean),
  rowMeans(data)
) %>% plot()
```

## sapply vs vapply

```{r}
bench::mark(
  sapply(mtcars, sum),
  vapply(mtcars, sum, FUN.VALUE = numeric(1))
) %>% plot()
```

## any all function

If you want to see if a vector contains a value, using `any(data == 10)` can be faster then `10 %in% data`. The speed will depend on the length of your vector and how frequent the value is.

```{r}
bench::press(
  nval = c(10, 100, 1000),
  {
    data <- rpois(nval, 1)
    bench::mark(
      5 %in% data,
      any(data == 5)
    )
  }
) %>% print() %>% plot()
```

## cut vs findInterval

Many times when you want to use `cut`, to split a vector into bins, it will be faster to replace cut with findInterval as it does less work by not creating a factor.

```{r}
x <- rnorm(10000)
bench::mark(check = FALSE,
  cut(x, c(-Inf, 0, Inf)),
  findInterval(x, c(-Inf, 0, Inf))
) %>%  print() %>% plot()
```

## unlist

`unlist` can be be speed up by setting `use.names = FALSE`, this will stop the returned object from being names, saving time.

```{r}
bench::mark(check = FALSE,
  unlist(mtcars),
  unlist(mtcars, use.names = FALSE)
)
```

## cumulative function and other math group functions

If you have any mathematical calculations you need to have done, it is worth your time to check if they are implemented in R already `?Math`.

## How many times does Y appear in X

Here we have some object which could be a vector, matrix or array. we would like to know how many times the number 0 appear.

```{r}
x_vector <- rpois(10000, 1)

bench::mark(
  unname(table(x_vector)["0"]),
  sum(x_vector == 0)
)
```

## table vs tabulate

If you need counts of integer-valued vectors, you will gain a big speedup by using the `tabulate()` function. Notice that tabulate will ignore non-positive integers, so 0s would not be counted. tabulate will manually place the number of counts of 8 in the 8th place in its output.

```{r}
x <- sample(1:10, 1000, TRUE)

table(x)
tabulate(x)

bench::mark(check = FALSE,
  table(x),
  tabulate(x)
)
```

## Using pipes

Use of the pipe operator `%>%` can make your code more readable, but it will also result in (slightly) slower code. If speed is very important you might want to reconsider using pipes. This will be a trade-off between readability and speed. 

```{r}
library(magrittr)
x <- rpois(1000, 1)

bench::mark(
  x %>% identity(),
  identity(x)
) %>% 
  print() %>% 
  plot()
```

## Summing weights

```{r}
x <- matrix(rnorm(1000), ncol = 1)
w <- matrix(rnorm(10000), ncol = 10)

bench::mark(
  t(x) %*% w,
  crossprod(x, w)
)
```

## pmax and pmin

pmin is a nice shorthand, but you can gain some speed with some rewriting.

```{r}
x <- runif(1000, -0.5)

bench::mark(
  pmax(x, 0),
  {
    i <- x < 0
    x[i] <- 0
    x
  }
)
```

## lm

If you are doing linear regression, and you already have the data as matrices you can benefit greatly from using the `.lm.fit` function. This function will not do any checking for you. Just the calculations.

```{r}

y <- mtcars$mpg

x <- as.matrix(mtcars[, c("disp", "cyl")])

bench::mark(check = FALSE,
  lm(mpg ~ disp + cyl - 1, mtcars),
  .lm.fit(x, y)
)

```

## using : for creating sequences

Creating a sequence from 1 to n.

```{r}
n <- 100000

bench::mark(
  seq(1, n),
  seq.int(1, n),
  1:n,
  seq_len(n)
)
```

Not only is `seq_len()` faster, but it is also safer as it as it works well when `n = 0`.

```{r}
1:0
seq_len(0)
```
