---
output: 
  slidy_presentation:
    theme: journal
    highlight: haddock
date: March 23, 2017
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, comment = "#")
```

## The `parallel` package

*   A refurbished version of `snow`

*   Shipped with R (part of R base)

*   Very flexible

## Parallel workflow

1.  Create a cluster:

    a.  PSOCK Cluster: `makePSOCKCluster`: Creates brand new R Sessions (so
        nothing is inherited from the master), even in other computers!
        
    b.  Fork Cluster: `makeForkCluster`: Using OS
        [Forking](https://en.wikipedia.org/wiki/Fork_(system_call)),
        copies the current R session locally (so everything is inherited from
        the master up to that point). Not available on Windows.
    
    c.  Other: `makeCluster` passed to **snow**
    
2.  Copy/prepare each R session:

    a.  Copy objects with `clusterExport`

    b.  Pass expressions with `clusterEvalQ`

    c.  Set a seed
    

3.  Do your call:

    a.  `mclapply`, `mcmapply` if you are using **Fork**

    b.  `parApply`, `parLapply`, etc. if you are using **PSOCK**

    
4.  Stop the cluster with `clusterStop`
    
## parallel example 1: Parallel RNG

```{r parallel-ex1, echo=TRUE}
# 1. CREATING A CLUSTER
library(parallel)
cl <- makePSOCKcluster(2)    

# 2. PREPARING THE CLUSTER
clusterSetRNGStream(cl, 123) # Equivalent to `set.seed(123)`

# 3. DO YOUR CALL
ans <- parSapply(cl, 1:2, function(x) runif(1e3))
(ans0 <- var(ans))

# I want to get the same!
clusterSetRNGStream(cl, 123)
ans1 <- var(parSapply(cl, 1:2, function(x) runif(1e3)))

ans0 - ans1 # A matrix of zeros

# 4. STOP THE CLUSTER
stopCluster(cl)
```

-----

In the case of `makeForkCluster`

```{r parallel-ex1-cont, echo=TRUE}
# 1. CREATING A CLUSTER
library(parallel)

# The fork cluster will copy the -nsims- object
nsims <- 1e3
cl    <- makeForkCluster(2)    

# 2. PREPARING THE CLUSTER
RNGkind("L'Ecuyer-CMRG")
set.seed(123) 

# 3. DO YOUR CALL
ans <- do.call(cbind, mclapply(1:2, function(x) {
  runif(nsims) # Look! we use the nsims object!
               # This would have fail in makePSOCKCluster
               # if we didn't copy -nsims- first.
  }))
(ans0 <- var(ans))

# Same sequence with same seed
set.seed(123) 
ans1 <- var(do.call(cbind, mclapply(1:2, function(x) runif(nsims))))

ans0 - ans1 # A matrix of zeros

# 4. STOP THE CLUSTER
stopCluster(cl)
```

## parallel example 2: Simulating $\pi$


*   We know that $\pi = \frac{A}{r^2}$. We approximate it by randomly adding
    points $x$ to a square of size 2 centered at the origin.

*   So, we approximate $\pi$ as $\Pr\{\|x\| \leq 1\}\times 2^2$

```{r, echo=FALSE, dev='jpeg', dev.args=list(quality=100), fig.width=6, fig.height=6, out.width='300px', out.height='300px'}
set.seed(1231)
p    <- matrix(runif(5e3*2, -1, 1), ncol=2)
pcol <- ifelse(sqrt(rowSums(p^2)) <= 1, adjustcolor("blue", .7), adjustcolor("gray", .7))
plot(p, col=pcol, pch=18)
```

The R code to do this

```{r simpi, echo=TRUE}
pisim <- function(i, nsim) {  # Notice we don't use the -i-
  # Random points
  ans  <- matrix(runif(nsim*2), ncol=2)
  
  # Distance to the origin
  ans  <- sqrt(rowSums(ans^2))
  
  # Estimated pi
  (sum(ans <= 1)*4)/nsim
}
```


-----

```{r parallel-ex2, echo=TRUE, cache=TRUE}

# Setup
cl <- makePSOCKcluster(10)
clusterSetRNGStream(cl, 123)

# Number of simulations we want each time to run
nsim <- 1e5

# We need to make -nsim- and -pisim- available to the
# cluster
clusterExport(cl, c("nsim", "pisim"))

# Benchmarking: parSapply and sapply will run this simulation
# a hundred times each, so at the end we have 1e5*100 points
# to approximate pi
rbenchmark::benchmark(
  parallel = parSapply(cl, 1:100, pisim, nsim=nsim),
  serial   = sapply(1:100, pisim, nsim=nsim), replications = 1
)[,1:4]

```

```{r printing-and-stop, cache=TRUE}
ans_par <- parSapply(cl, 1:100, pisim, nsim=nsim)
ans_ser <- sapply(1:100, pisim, nsim=nsim)
stopCluster(cl)
```

```{r, echo=FALSE}
c(par = mean(ans_par), ser = mean(ans_ser), R = pi)
```

## RcppArmadillo and OpenMP

*   Must run only 'Thread-safe' calls, so calling R within parallel blocks can cause
    problems (almost all the time).
    
*   Use `arma` objects, e.g. `arma::mat`, `arma::vec`, etc. Or, if you are used to them
    `std::vector` objects as these are thread safe.

*   Pseudo Random Number Generation is not very straight forward.

*   

*   Data directives: Tell the compiler how to treat data. (By default `shared`).
    
    *   `shared`: Default, all threads access the same copy.
    *   `private`: Each thread has its own copy (although not initialized).
    *   `firstprivate` Each thread has its own copy initialized.
    *   `lastprivate` Each thread has its own copy. The last value is the one stored in the main program.

## RcppArmadillo + OpenMP example 1: Distance matrix

```cpp
#include <omp.h>
#include <RcppArmadillo.h>

// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::plugins(openmp)]]

using namespace Rcpp;

// [[Rcpp::export]]
arma::mat dist_par(arma::mat X, int cores = 1) {
  // Some constants
  int N = (int) X.n_rows;
  int K = (int) X.n_cols;
  
  // Output
  arma::mat D(N,N);
  D.zeros(); // Filling with zeros
  
  // Setting the cores
  omp_set_num_threads(cores);
  
#pragma omp parallel for shared(D, N, K, X) default(none)
  for (int i=0; i<N; i++)
    for (int j=0; j<i; j++) {
      for (int k=0; k<K; k++) 
        D.at(i,j) += pow(X.at(i,k) - X.at(j,k), 2.0);
      
      // Computing square root
      D.at(i,j) = sqrt(D.at(i,j));
      D.at(j,i) = D.at(i,j);
    }
      
  
  // My nice distance matrix
  return D;
}
```

----

Let's see how it goes!

```{r openmp-example1, cache=TRUE, echo=TRUE}
# Compiling the function
Rcpp::sourceCpp("dist.cpp")

# Simulating data
set.seed(1231)
K <- 5000
n <- 500
x <- matrix(rnorm(n*K), ncol=K)

# Are we getting the same?
table(as.matrix(dist(x)) - dist_par(x, 10)) # Only zeros

# Benchmarking!
rbenchmark::benchmark(
  dist(x),                 # stats::dist
  dist_par(x, cores = 1),  # 1 core
  dist_par(x, cores = 4),  # 4 cores
  dist_par(x, cores = 10), # 10 cores
  replications = 1, order="elapsed"
)[,1:4]
```


## RcppArmadillo + OpenMP example 2: Simulating $\pi$

```cpp
#include <omp.h>
#include <RcppArmadillo.h>
using namespace Rcpp;

// [[Rcpp::depends(RcppArmadillo)]]
// [[Rcpp::plugins(openmp)]]

// [[Rcpp::export]]
double sim_pi(int m, int cores = 1) {
  
  // Setting the cores
  omp_set_num_threads(cores);
  int n = m / cores;
  
  double ans = 0.0, d;
  double val = 4.0/m;
  double piest;
  int i;
  
  // Pseudo RNG is not easy in OMP
  arma::mat points(m, 2);
  for (i = 0;i< (int) points.n_rows; i++)
    points.at(i, 0) = unif_rand()*2.0 - 1,
      points.at(i, 1) = unif_rand()*2.0 - 1;
  
#pragma omp parallel default(none) shared(ans, cores, points) \
  firstprivate(val, n, m) \
  private(piest, i, d)
{
    
  // Which core are we
  int core_num = omp_get_thread_num();
    
  piest = 0.0;
  for (i = n*core_num; i < (n + n*core_num); i++) {
    
    d = sqrt(pow(points.at(i, 0), 2.0) + pow(points.at(i, 1), 2.0));
    
    if (d <= 1.0)
      piest += val;
  }
  
  // This bit of code is executed one thread at a time.
  // Instead of -atomic-, we could have use -critical-, but that has
  // a higher overhead.
#pragma omp atomic
  ans += piest;
}
  
  return ans;
}
```

------

```{r pi-cpp, cache=TRUE, echo=TRUE}
# Compiling c++
Rcpp::sourceCpp("simpi.cpp")

# Running in 1 or 10 cores should be the same
set.seed(1); sim_pi(1e5, 1)
set.seed(1); sim_pi(1e5, 10)

# Benchmarking
nsim <- 1e8
benchmark(
  pi01 = sim_pi(nsim, 1),
  pi04 = sim_pi(nsim, 4),
  pi10 = sim_pi(nsim, 10),
  replications = 1
)[,1:4]

```

## Some thoughts on Rcpp + OpenMP

*   Setting `default(none)` is a good practice. The programmer must define what's shared and what's not.

*   Avoid **false sharing**: This happens when independent variables are written in the
    same cache line, so threads make each other slow (what?). In other words, Thread A
    is writing on element X[1] at the same time that Thread B wants to read X[2],
    but X[1] and X[2] are in the same cache line, so A 'blocks' the line and forces B to create
    a new copy of X[2], even when X[2] is not been used!
    
## Thanks!

```{r session, echo=FALSE}
devtools::session_info()
```



