---
title: "Monte Carlo Integration"
author: "George"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monte Carlo Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Definition

From Wolfram MathWorld:

> Picking $N$ randomly distributed points $x_1, x_2, ..., x_N$ in a multidimensional volume $V$ to determine the integral of a function $f$ in this volume gives a result

$$
\tag{1}
\int f dV \approx V \langle f\rangle \pm V\sqrt{\frac{\langle f^2 \rangle- \langle f \rangle^2}{N}}
$$
Where 

$$
\begin{align}
\langle f \rangle & = \frac{1}{N}\sum^N{i=1}f(x_i)\tag{2}\\
\langle f \rangle & = \frac{1}{N}\sum^N{i=1}f(x_i)^2\tag{3}
\end{align}
$$

# Example with 2D Multivariate Normal

Integrate between a = {-1, -10} and b = {1, 1}

```{r}
set.seed(1231)
library(mvtnorm)
library(numint)

ans <- num_int(
  dmvnorm, a = c(-1, -10), b = c(1, 1),
  mean = c(0,0), sigma = diag(2),
  N = 1e7, ncores = 2
)
ans$val
pmvnorm(lower = c(-1,-10), upper = c(1,1), mean = c(0,0), sigma = diag(2))
```

We can visualize a single dim of the area

```{r}
plot(ans)
```



# References

Weisstein, Eric W. "Monte Carlo Integration." From _MathWorld_--A Wolfram Web Resource. http://mathworld.wolfram.com/MonteCarloIntegration.html