---
title: "Workshop: Hands-on on Writing R packages"
author: 'George G. Vega Yon<br><a href="http://cana.usc.edu/vegayon">cana.usc.edu/vegayon</a><br><a href="mailto:vegayon@usc.edu">vegayon@usc.edu</a>'
date: "June 19, 2017"
output: 
  slidy_presentation:
    footer: vegayon@usc.edu
    incremental: false
    font_adjustment: -1
    duration: 45
    # css: ggvy_style.css
    slide_level: 2
    highlight: haddock
---

```{r, echo = FALSE}
knitr::opts_chunk$set(out.width = "600px", fig.align = "center")
```

# Introduction and General Overview

**Why writing an R package?**

To name a few:

1.  Easy way to share code: Just type `install.packages` and voilà!

2.  Already standarized: So you don't need to think about how to structure it.

3.  CRAN checks everything for you: Force yourself to code things right.

4.  **Reproducible Research**
    
    *   Accessible Reproducible Research ([Mesirov, __Science__ 2010](http://science.sciencemag.org/content/327/5964/415))
    
    *   Again, and Again, and Again, ... ([Jasny et al., __Science__ 2011](http://science.sciencemag.org/content/334/6060/1225))
        
    *   Challenges in Irreproducible Research ([__nature__ topic](http://www.nature.com/news/reproducibility-1.17552))


----

**What's an R package, anyway? (cont.)**

According to Hadley Wickham's ["R packages"](http://r-pkgs.had.co.nz/)

> Packages are **the fundamental units of reproducible R code**. They include
**reusable R functions**, the **documentation** that describes how to use them, and
**sample data**.

```{r What it has, echo=FALSE}
words <- c("R/", "tests/","vignettes/", "data/", "man/",
           "DESCRIPTION", "NAMESPACE")
freqs <- 7:1
# oldpar <- par(no.readonly = TRUE)
# par(mar = rep(0,4))
# wordcloud::wordcloud(words, colors = c(blues9[5:9], gray(c(.8, .6)))[7:1],
#                      freqs, min.freq = 1)
# par(oldpar)

# devtools::install_github("jbkunst/d3wordcloud")
library(d3wordcloud)
d3wordcloud(words, freqs, colors = c(blues9[5:9], gray(c(.8, .6))))
```

----

**What's an R package, anyway? (cont.)**

Folders

*   **R** Where the R scripts live, e.g. `addnums.r`, `funny-pkg.r`.

*   **tests** Where the tests live, e.g. `testthat/test-addnums.r` (using testthat).

*   **vignettes** Where vignettes live, e.g. `vignettes/mode_details_on_addnums.rmd`.

*   **data** Where the data lives, e.g. `fakedata.rda`, `aneattable.csv`.

*   **man** Where the manuals live, e.g. `addnums.Rd`, `funnypkg.Rd`.

And files

*   **DESCRIPTION** The metadata of the package: name, description, version, author(s),
    dependencies, etc.

*   **NAMESPACE** Declares the name of the objects (functions mostly) that will
    be part of the package (and will be visible to the user)

More info [here](http://r-pkgs.had.co.nz/package.html) and [here](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Package-structure)


----

Writing R packages is an iterative process

1.  **Set up the structure**: Create folders and files `R/`, `data/`, `src/`, `tests/`, `man/`,
    `vignettes/`, `DESCRIPTION`, `NAMESPACE`(?)

2.  **Code!** For `f` in `F` do
    
    1.  Write `f`
    
    2.  Document `f`: what it does, what the inputs are, what it returns, examples.
    
    3.  Add some tests: is it doing what is supposed to do?
    
    4.  `R CMD Check`: Will CRAN, and the rest of the functions, 'like' it?
    
    5.  Commit your changes!: Update `ChangeLog`+`news.md`, and commit
        (so travis and friends run)!
    
    6.  next `f`
    
3.  Submit your package to CRAN/Bioconductor (read [more](http://r-pkgs.had.co.nz/release.html#release-submission))

----

**How are we going to work today**

1.  Each "topic" will be introduced in the general context of CRAN's requirements.

2.  Explain how the CRAN requirment works by examples.

3.  Pick one or more volunteers so we can ilustrate the presentation with their projects.

4.  Actually work you the projects... :).


# Lightning talks: Presentation of the Projects


----

<h1>
<center>15 min Break</center><br>
</h1>

----


# Setting Up the Rstudio+git project

Create an R Package project

*   Go to "File" > "New Project" and and select "R Package" in project type

*   RStudio will create a new R package project with a template set of functions and files that you will be able to modify (as the ones we described in the introduction).


# Setting Up the Rstudio+git project (cont. 1)

Using the `devtools` package, we can add some extras

```r
# Creating a README for the project
devtools::use_readme_rmd()

# Infrastructure for testing
devtools::use_testthat()

# Infrastructure for Code Coverage
devtools::use_coverage(type = "codecov") # This creates the .travis.yml
devtools::use_appveyor()

# A LICENSE file (required by CRAN)
# The line "license: MIT + file" in the DESCRIPTION file
devtools::use_mit_license(copyright_holder = "George G. Vega Yon")
devtools::use_news_md()
```

# Setting Up the Rstudio+git project (cont. 2)

What did just happen?

1.  `use_readme_rmd()`: Creates a readme file that will be in the main folder
    of the project. Think about it as the home page of your project.
    
2.  `use_testthat()`: Will create the basic infrastructure for package testing.

3.  `use_coverage(type = "codecov")`: Creates the [.travis.yml file](https://en.wikipedia.org/wiki/Comparison_of_continuous_integration_software)
    for Unix CI, and
    sets it up for code coverage using [codecov.io](https://codecov.io/).
    
4.  `devtools::use_appveyor()`: Creates the [appveyor.yml](https://www.appveyor.com/)
    for windows CI.
    
5.  `devtools::use_mit_license(copyright_holder = "George G. Vega Yon")`: Creates the----
    `LICENSE` file and puts it under 'George G. Vega Yon'.

6.  `devtools::use_news_md()`: Creates the `news.md` file which us used for
    tracking changes, and communicating them to the users (e.g.
    [netdiffuseR](https://github.com/USCCANA/netdiffuseR/blob/master/NEWS.md))

# Setting Up the Rstudio+git project (cont. 3)

For Git, you will need to create a project on github. Just go to your account:

*   Go to http://github.com and log in into your account to create a new repository.
    If you are already logged in, you can simple go to http://github.com/new

*   You'll need to initialize the repository with data. For that we all all the files (source code file) tha should be tracked by git and/or need to be included so that the package can be built.

*   To add a file to your repository just type
    
    ```shell
    $ git add [list of files to add]
    ```
    
    To remove a file (only from the tracking), type
    
    ```shell
    $ git rm --cached [list of files to remove]
    ```
    
*   To make a commit, just type
    
    ```shell
    $ git commit -a m "[Comment about the commit]"
    ```

*   For the first commit, you need to (1) add the remote repo, and (2) include the -u flag  
    
    ```shell
    $ git remote add origin https://github.com/gvegayon/numint.git
    $ git push -u origin master
    ```
    
----

Once you have your first commit, the (git)dev cycle is pretty much as follows:
    
1.  Pull the changes 
    
    ```shell
    $ git pull
    ```
    
2.  Edit/add/remove code/files.

3.  Add files (if any) 
    
    ```shell
    $ git add [list of files to add]
    ```

4.  Commit the changes 
    
    ```shell
    $ git commit -a -m "[a brief comment on what you did]"
    ```

5.  Push your changes 
    
    ```shell
    $ git push
    ```

**Obs** Besides of the comment that you add to the commit, it is recommended to include a [`ChangeLog`](https://en.wikipedia.org/wiki/Changelog)
Examples: https://github.com/USCCANA/netdiffuseR/blob/master/ChangeLog and https://github.com/USCbiostats/aphylo/blob/master/ChangeLog

----

<h1>
<center>Lets see some examples!</center>
</h1>

# Documenting an R package with roxygen2: functions

*   Using `roxygen2` is very straight forward. For our fist pice of code, we create the
    `R/addnums.r` file:
        
    ```{r}
    #' The title of -addnums-
    #'
    #' Here is a brief description
    #' 
    #' @param a Numeric scalar. A brief description.
    #' @param b Numeric scalar. A brief description.
    #' 
    #' @details Computes the sum of \code{x} and \code{y}.
    #' @return A list of class \code{funnypkg_addnums}:
    #' \item{a}{Numeric scalar.}
    #' \item{b}{Numeric scalar.}
    #' \item{ab}{Numeric scalar. the sum of \code{a} and \code{b}}
    #' @examples
    #' foo(1, 2)
    #' 
    #' @export
    addnums <- function(a, b) {
      ans <- a + b
      structure(list(a = a, b = b, ab = ans)
      , class = "funnypkg_addnums")
    }
    
    ```

*   Type `devtools::document()`, or press Ctrl + Shift + D (RStudio will: create the manual, and the NAMESPACE).
    Make sure you activate this option in RStudio (not the default).
    
*   Notice that the output is defined using S3-type objects ([read more](http://adv-r.had.co.nz/S3.html))

---


*   Here we are adding a function that is documented in `addnums`. The plot method
    
    ```{r}
    #' @rdname addnums
    #' @export
    #' @param x An object of class \code{funnypkg_addnums}.
    #' @param y Ignored.
    #' @param ... Further arguments passed to
    #' \code{\link[graphics:plot.window]{plot.window}}.
    plot.funnypkg_addnums <- function(x, y = NULL, ...) {
      graphics::plot.new()
      graphics::plot.window(xlim = range(unlist(c(0,x))), ylim = c(-.5,1))
      graphics::axis(1)
      with(x, graphics::segments(0, 1, ab, col = "blue", lwd=3))
      with(x, graphics::segments(0, 0, a, col = "green", lwd=3))
      with(x, graphics::segments(a, .5, a + b, col = "red", lwd=3))
      graphics::legend("bottom", col = c("blue", "green", "red"), 
             legend = c("a+b", "a", "b"), bty = "n", 
             ncol = 3, lty = 1, lwd=3)
    }
    
    ```
      
*   You'll need to update the `man/*Rd` every time that you add new roxygen content.
    Just press Ctrl + Shift + D and RStudio will do it for you.
    
*   Notice that functions from foreign packages are called using the `::` operator
    (read more [here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/ns-dblcolon.html)
    and [here](http://r-pkgs.had.co.nz/namespace.html)). Not a requirement, but
    makes the code easier to maintain (and less error-prone... because of R's scoping rules).

---

Since we added foreign functions, we need to add them to the NAMESPACE and
to the DESCRIPTION files:

*   Here is the `R/funny-pkg.r` file:
    
    ```r
    #' @importFrom graphics plot.new plot.window axis legend
    NULL
    
    #' funnypkg
    #'
    #' A (not so) funny collection of functions
    #'
    #' @description We add stuff up... You can access to the project
    #' website at \url{https://github.com/USCbiostats/software-dev/tree/master/rpkgs/funnypkg}
    #'
    #' @docType package
    #' @name funnypkg
    #'
    #' @author George G. Vega Yon
    NULL
    ```
    
*   Here is the DESCRIPTION updated
    
    ```
    Package: funnypkg
    Type: Package
    Title: What the Package Does (Title Case)
    Version: 0.1.0
    Author: Who wrote it
    Maintainer: The package maintainer <yourself@somewhere.net>
    Description: More about what it does (maybe more than one line)
        Use four spaces when indenting paragraphs within the Description.
    License: What license is it under?
    Encoding: UTF-8
    LazyData: true
    Suggests:
        testthat,
        covr
    RoxygenNote: 5.0.1
    Imports: graphics
    ```
    
*   The `Suggests` and `RoxygenNote` fields where added automagically by `devtools`.


----

<h1>
<center>15 min Break</center><br>
</h1>


----

# Documenting an R package with roxygen2: Dependencies, data and the package itself

```r
#' netdiffuseR
#'
#' Analysis of Diffusion and Contagion Processes on Networks
#'
#' @description
#'
#' Statistical analysis, visualization and simulation of diffusion and contagion
#' processes on networks. The package implements algorithms for
#' calculating stats such as innovation threshold levels, infectiousness
#' (contagion) and susceptibility, and hazard rates as presented in Burt (1987),
#' Valente (1995), and Myers (2000) (among others).
#'
#' You can access to the project website at
#' \url{https://github.com/USCCANA/netdiffuseR}
#'
#' @section Acknowledgements:
#' netdiffuseR was created with the support of grant R01 CA157577 from the National Cancer Institute/National Institutes of Health.
#'
#' @docType package
#' @name netdiffuseR
#'
#' @author Vega Yon, Dyal, Hayes & Valente
#'
NULL
```

----

Documenting *Dependencies* from the [netdiffuseR](https://github.com/USCCANA/netdiffuseR/blob/master/R/imports.R) R package

```r
#' @importFrom Rcpp evalCpp
NULL

#' @importFrom sna gplot as.sociomatrix.sna
NULL

#' @importFrom igraph graph_from_adjacency_matrix set_vertex_attr
#'  any_multiple graph_attr_names as_adj is.loop set_graph_attr V permute
#' @importFrom network as.edgelist is.multiplex is.directed has.loops
NULL

#' @useDynLib netdiffuseR
NULL

# Importing from the Matrix pkg ------------------------------------------------

#' @import methods
#' @importClassesFrom Matrix dgCMatrix
#' @importFrom Matrix Matrix rowSums colSums
#' @importMethodsFrom Matrix t
NULL

#' @importClassesFrom SparseM matrix.csc
NULL

# Importing from R CORE packages -----------------------------------------------

#' @importFrom grDevices grey rgb colorRampPalette blues9
#' @importFrom graphics grid par plot points symbols text layout legend lines
#'  matplot plot.new plot.window hist mtext polygon image title .filled.contour
#' @importFrom stats complete.cases runif reshape setNames ftable sd pnorm var
#' as.formula optim nls coef
#' @importFrom utils getFromNamespace head str
#' @importFrom boot boot
#' @importFrom MASS bandwidth.nrd kde2d
#' @importFrom MatchIt matchit
NULL

release_questions <- function() {
  c(
    "Have you updated the inst/NEWS file?",
    "Have you changed the version+dates in DESCRIPTION and NEWS.md?",
    "Have you added all new files to GIT?",
    "Have you clean the vignettes file (source)?"
    )
}
```

Which genertes this [NAMESPACE](https://github.com/USCCANA/netdiffuseR/blob/master/NAMESPACE)

-----

An example of data documentation can be found in the R package [`aphylo`](https://github.com/USCbiostats/aphylo/blob/master/R/package.r#L56-L86)


```r
#' Fake Phylogenetic Tree
#' 
#' A fake dataset containing the parent-offspring relations between genes.
#' This dataset is inteded for testing only.
#' 
#' @format A data frame with 6 rows and 2 variables:
#' \describe{
#'   \item{NodeId}{Integer, ID of the offspring.}
#'   \item{ParentId}{Integer, ID of the parent.}
#' }
#' 
#' @source BiostatsUSC
#' @name faketree
NULL

#' Fake Experimental Data
#' 
#' A fake dataset containing 2 functional state of the leaf nodes. Each
#' function can have either 0 (unactive), 1 (active) or 9 (n/a).
#' This dataset is inteded for testing only.
#' 
#' @format A data frame with 4 rows and 3 variables:
#' \describe{
#'   \item{f1}{State of function 1.}
#'   \item{f2}{State of function 1.}
#'   \item{LeafId}{Integer, ID of the leaf.}
#' }
#' 
#' @source BiostatsUSC
#' @name fakeexperiment
NULL
```


# Writing tests with testthat

> Software testing is important, but, in part because it is frustrating and boring, many of us avoid it.
> 
> testthat is a new testing framework for R that is easy learn and use, and integrates with your existing workflow. (?testthat)

----

Test are run every time that you run R CMD check

*   In the `tests/testthat/` dir, add/edit a source file with tests, e.g. `test-addnums.r`
    
    ```r
    context("Basic set of tests")
    test_that("addnums(a, b) = a+b", {
      # Preparing the test
      a <- 1
      b <- -2
      
      # Calling the function
      ans0 <- a+b
      ans1 <- addnums(a, b)
      
      # Are these equal?
      expect_equal(ans0, ans1$ab)
    })
    
    test_that("Plot returns -funnypkg_foo-", {
      expect_s3_class(plot(addnums(1,2)), "funnypkg_foo")
    })
    ```
    
*   You can run the tests using Ctrl + Shift + t

*   Furthermore, tests are use to evaluate code coverage.


-----------

<h1>
<center>15 min Break</center><br>
</h1>

# Ready to publish: R CMD check

Running `R CMD check` is easy with RStudio

*   If you don't have C/C++  code Just press Ctrl + Shift + E

*   If you have C/C++ code, use `R CMD Check` with valgrind (check for segfaults)
    
    ```shell
    $ R CMD build funnypkg/
    $ R CMD check --as-cran --use-valgrind funnypkg*.tar.gz
    ```
    
    You can ask RStudio to use valgrind too.

For an extensive list of the checks, see the section
[1.3.1 Checking packages](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Checking-packages)
in the "Writing R extensions" manual.


-----

Once you have run R CMD checks and no errors or unexpected issues are detected,
you should commit your changes and push them to Github.

1.  Edit the [`NEWS.md`](https://github.com/hadley/dplyr/blob/master/NEWS.md) file and the
    [`ChangeLog`](https://github.com/RcppCore/RcppArmadillo/blob/master/ChangeLog) (if you have one).

2.  Add the new files to the tree, commit your changes, and push them to Github
    
    ```shell
    $ git add R/newfile.r man/newfile.Rd
    $ git commit -a -m "Adding new function"
    $ git push
    ```
    
Once uploaded, I recommend checking Travis and AppVeyor to see if everything
went well.

-----

*   You can release your package using `devtools::release()` function.

*   The devtools package was built to make life easier. Such is it's creator's
    confidence on the `release` function that it comes with a warranty!
    
    > If a devtools bug causes one of the CRAN maintainers to **treat you
      impolitely**, I will personally send you a handwritten apology note. Please
      forward me the email and your address, and I'll get a card in the mail.

# Dev Cycle (once again!)

The most important step

2.  **Code!** For `f` in `F` do
    
    1.  Write `f`
    
    2.  Document `f`: what it does, what the inputs are, what it returns, examples.
    
    3.  Add some tests: is it doing what is supposed to do?
    
    4.  `R CMD Check`: Will CRAN, and the rest of the functions, 'like' it?
    
    5.  Commit your changes!: Update `ChangeLog`+`news.md`, and commit
        (so travis and friends run)!
    
    6.  next `f`
  

# Beyond examples: Writing Vignettes

What are vignettes? From the "Writing R Extensions" manual, [section 1.4](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Writing-package-vignettes)

> Vignettes are documents in PDF or HTML format obtained from plain text literate source files from which R knows how to extract R code and create output (in PDF/HTML or intermediate LaTeX).

> Vignette engines do this work, using “tangle” and “weave” functions respectively. Sweave, provided by the R distribution, is the default engine. Since R version 3.0.0, other vignette engines besides Sweave are supported; see Non-Sweave vignettes.

----

An example vignette for the [numint R package](http://github.com/gvegayon/numint)

```{r, results='markup', comment="", warning=FALSE}
cat(paste(readLines("numint.rmd"), collapse="\n"), fill = TRUE)
```


# Workshop: Hands-on on Writing R packages

<body>
<p align="center" style="font-size:50px"><b>Thanks!</b></pr><br>
<p align="center">George G. Vega Yon<br><a href="http://cana.usc.edu/vegayon">cana.usc.edu/vegayon</a><br><a href="mailto:vegayon@usc.edu">vegayon@usc.edu</a></p>
</body>

## More resources

*   ["Writing R extensions"](https://cran.r-project.org/doc/manuals/r-release/R-exts.html)
    (R Core Team)

*   ["R packages"](http://r-pkgs.had.co.nz/) (Hadley Wickham)

*   ["Developing Packages with RStudio"](https://support.rstudio.com/hc/en-us/articles/200486488-Developing-Packages-with-RStudio)
    (Josh Paulson)
    
*   ["A template for R packages using Rcpp and/or RcppArmadillo "](https://github.com/patrickmuchmore/RPackageTemplate)
    (Patrick Muchmore)

*   ["R package primer: a minimal tutorial"](http://kbroman.org/pkg_primer/) (Karl Broman)

*   ["Writing an R package from scratch"](https://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/)
    (Hilary Parker)

*   ["Instructions for Creating Your Own R Package"](http://web.mit.edu/insong/www/pdf/rpackage_instructions.pdf)
    (Kim _et al._)
